Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-22 submission/manuscript.Rmd
add third results section

**Random Forest models with SCFA concentrations do not classify tumor better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations. For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although there was no added information gained for classification of disease by including SCFAs to Random Forest models, it is possible that the genes for enzymes involved in SCFA pathways may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.

>>>>

**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations. For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although there was no added information gained for classification of disease by including SCFAs to Random Forest models, it is possible that the genes for enzymes involved in SCFA pathways may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. We next took a subset of these `r length(metai$sample)` fecal samples and performed metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons and patients with adenoma or carcinoma (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. This lack of difference can be highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to tumors. Overall, there may be some systematic bias present within our results that may account for the lack of associations.

### 2018-05-22 submission/manuscript.tex
add third results section

\textbf{Random Forest models with SCFA concentrations do not classify
tumor better.} Using the Random Forest algorithm we built models to
classify normal versus adenoma and normal versus carcinoma with OTUs or
OTUs and SCFA concentrations. For both adenoma and carcinoma models,
there was no difference between the median AUC of models with or without
SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
Although there was no added information gained for classification of
disease by including SCFAs to Random Forest models, it is possible that
the genes for enzymes involved in SCFA pathways may change due to
disease status; where a smaller number of microbes are responsible for
the observed SCFA concentrations.

>>>>

\textbf{Random Forest models with SCFA concentrations do not classify
tumors better.} Using the Random Forest algorithm we built models to
classify normal versus adenoma and normal versus carcinoma with OTUs or
OTUs and SCFA concentrations. For both adenoma and carcinoma models,
there was no difference between the median AUC of models with or without
SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
Although there was no added information gained for classification of
disease by including SCFAs to Random Forest models, it is possible that
the genes for enzymes involved in SCFA pathways may change due to
disease status; where a smaller number of microbes are responsible for
the observed SCFA concentrations.

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes inolved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. We next took a subset of these 490 fecal samples and
performed metagenomic sequencing to confirm these results. Like the
imputed gene results, metagenomic analysis found that there was no
difference in any of the genes involved in SCFA synthesis between
individuals with normal colons and patients with adenoma or carcinoma
(P-value \textgreater{} 0.70) {[}Table S3{]}. This lack of difference
can be highlighted when we visualize the results for butyrate kinase
{[}Figure 3B{]}. These observations provide evidence that the gene
content does not change due to tumors. Overall, there may be some
systematic bias present within our results that may account for the lack
of associations.

### 2018-05-22 submission/supplemental.Rmd
add third results section

>>>>

\newpage

**Table S2: Breakdown of Kruskal-Wallis tests using PICRUSt imputed relative abundance of select genes in SCFA pathways.**
```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
table_s2 <- read_csv("../data/process/specific_scfa_kruskal_picrust_summary.csv") %>% 
  select(kegg_ortholog, statistic, p.value, bh) %>% 
  mutate_at(vars(statistic:bh), function(x) round(x, digits = 3))

colnames(table_s2) <-  c("KEGG IDs", "Rank Sum Statistic", "P-value", "BH")

kable(table_s2, format = "markdown", align = 'c', padding = 2)
# opf_select <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv")
```

\newpage

**Table S3: Breakdown of Kruskal-Wallis tests using counts per million from metagenomic sequencing of select genes in SCFA pathways.**
```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
table_s3 <- read_csv("../data/process/select_scfa_opf_kruskal_summary.csv") %>% 
  select(kegg_id, statistic, p.value, bh) %>% 
  mutate_at(vars(statistic:bh), function(x) round(x, digits = 3))

colnames(table_s3) <-  c("KEGG IDs", "Rank Sum Statistic", "P-value", "BH")

kable(table_s3, format = "markdown", align = 'c', padding = 2)
```
